Welcome Home --- Now Vote!

### 2020-03-05 markdown/part2/part2_standalone.Rmd
write

---
title: "Welcome Home --- Now Vote!"
subtitle: "Voting Rights Restoration and Post-Supervision Participation"
thanks: Prepared for the 2020 Annual Meeting of the Midwest Political Science Association. The author thanks Jacob Faber, Jeff Manza, Myrna Pérez, Ariel White, and Peter Miller for their comments on this project. All errors are my responsibility.
author: Kevin Morris^[Researcher, Brennan Center for Justice at NYU School of Law, 120 Broadway Ste 1750, New York, NY 10271 (kevin.morris@nyu.edu)]
institute: "Brennan Center for Justice"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "../welcome_home_vote.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
    - \usepackage{setspace}
abstract: |
  This paper presents causal estimates of the effect of voting rights restoration prior to discharge from parole on post-supervision participation. In 2018, New York State began restoring voting rights to parolees, after previously restoring voting rights only at the completion of parole. By leveraging randomness in parole discharge date, I interrogate whether restoring voting rights to parolees increases their post-supervision propensity to cast a ballot. I demonstrate that individuals whose rights were restored while still on parole were more likely to participate than those whose rights were automatically restored upon completion of parole. This group-level effect, however, masks race-specific effects. Although rights restoration prior to parole discharge effectively doubled turnout among white former parolees, it had no measurable effect on the turnout of non-white former parolees. This raises serious questions about how rights restoration programs are implemented, and how incarceration might differently structure Black Americans’ view of the democratic process.
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\pagebreak
\doublespacing
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "../..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(data.table)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

In all but two states (Maine and Vermont), felony disenfranchisement laws mean that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. In other states, notably Kentucky and Iowa, felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor [@bcj_laws]. This variation in laws arises from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights "for participation in rebellion, or other crime." The definition of "other crime," left so vague in the Constitution, is now generally used by states to disenfranchise citizens for any felony offense. The Supreme Court, in cases such as *Richardson v. Ramirez* (1974), has upheld this practice. Collectively, these laws disenfranchise as many as 4.7 million American citizens. Of these, the majority are no longer incarcerated[@sentencing_2016].^[The figures reported in @sentencing_2016 have been adjusted to reflect the impact of Amendment 4 in Florida.]

There is some evidence that incarceration continues to structure political participation even after an individual is no longer legally disenfranchised. As previous literature has established, interactions with the criminal justice system leaves residents less likely to vote in the future [@White2019; but see @Gerber2017]. As @Burch2011 and others have shown, moreover, turnout rates among the formerly incarcerated are extremely low. Formal disenfranchisement policy, the literature has made clear, is just one piece of an interlocking system that serves to disenfranchise minority and marginalized voters. The incarcerated population is drawn from a pool of individuals unlikely to vote even prior to their incarceration. To address only the formal laws contributing to disenfranchisement without also interrogating efforts to boost post-supervision participation risks leaving much of the system of effective disenfranchisement undisturbed. New York State offers us the opportunity to test how the timing of the re-instation of voting rights structures post-supervision participation.

@@

In the spring of 2018,^[Although the executive order was signed on April 18^th^, it did not actually go into effect until May 18^th^.] Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. Such a move was of course good for individuals who were still on parole on election day, and would have been disenfranchised otherwise. The change in policy is also beneficial for felony probationers: despite the fact that probationers do not formally lose their voting rights, there is evidence that confusion around the law contributes to *de facto* disenfranchisement among probationers [@Drucker2005]. The executive order is a promising step: by changing the policy to allow all New York citizens living in their communities to cast a ballot, the move has the potential to both re-enfranchise the nearly 30,000 New Yorkers on parole living in the community and to clarify the rules about who is eligible to vote.

There is reason to believe that the executive order may also have increased the political participation of formerly disenfranchised individuals. Prior to the policy change, formerly incarcerated individuals had their voting rights restored automatically upon the completion of their parole term. New York's correction code provides no more guidance other than that "upon a person's discharge from community supervision, the department shall notify such person of his or her right to vote and provide such person with a form of application for voter registration" (Section 75 of the New York State Correction Law). Shortly after the implementation of the executive order, Acting Deputy Commissioner for Community Supervision Ana Enright sent a memorandum to New York State parole officers detailing the Department of Corrections' new approach.^[Link: https://nyassembly.gov/member_files/139/webdocs/82103.pdf] The memorandum directs all parole officers to present parolees with voter registration forms and to explain their purpose. Parole officers were also instructed to offer any assistance needed, including help filling out the registration form. In addition to these directives, the memorandum communicates that voter registration is to receive "high priority attention," and it separately calls the program "a **priority** initiative" [emphasis in the original]. Thus, the executive order demands not only that re-enfranchised individuals receive in-person notification of their voting rights, but also that parole officers prioritize their registration.

In the analysis below, I examine the effect of Executive Order 181 on individuals who finished parole before October 10^th^, 2018 (the registration deadline for 2018). These individuals, because they were discharged from parole before the registration deadline, would have been eligible to vote even in the absence of the executive order. This is only a subset of the individuals who were impacted by Executive Order 181 --- many of the individuals who had their voting rights restored were still on parole on election day, and therefore were eligible to vote only because of the rules change. However, by focusing on individuals who would have been eligible to vote either way, I can test whether rights restoration prior to parole discharge serves as an effective encouragement of post-supervision participation.

By examining the effect of rights restoration in the context of a personal relationship with a parole officer, rather than simply at the end of a sentence and communicated through the mail, Executive Order 181 offers us the opportunity to build on existing research. Do these human interactions repair damage done to the formerly incarcerated individual's relationship to the state? Or do parole officers exhibit biases against their stewards, making them ineffective conduits for restoration?

@@

Much of this literature has established that formerly incarcerated individuals rarely vote, even when they are no longer formally barred from doing so. The causal effect of incarceration on participation is the subject of some debate within the field. Individuals who go to prison share many characteristics with lower propensity voters generally. Less educated citizens, for instance, turnout at low rates whether they have been to prison or not. In an attempt to disentangle sociodemographic characteristics from the experience of imprisonment, @Gerber2017 uses administrative data from Pennsylvania to estimate turnout rates prior to and after incarceration. They argue that the majority of the low turnout observed among formerly incarcerated individuals can be explained by observable characteristics, concluding that "it appears that spending time in prison does not have large negative effects on subsequent participation" (1144).

@@

Regardless of the precise mechanism, the low turnout among formerly incarcerated individuals is cause for concern, particularly given the racialized aspects of the criminal justice system. A criminal justice system that inflicts serious consequences on a population that has relatively little political voice is problematic. Whether or not incarceration *causes* low turnout the state has a unique opportunity to craft policies that will impact individuals under formal supervision. Even if incarceration does not lead to lower turnout, policies targeting individuals caught up in the criminal justice system might still be effective at increasing turnout for this population.

Formerly convicted individuals are very often confused about their eligibility to vote [@Drucker2005; @locked_out]. Some research indicates that dispelling misinformation boosts turnout among the formerly disenfranchised. @Meredith2015 examines the impact of ending permanent disenfranchisement in Iowa. Prior to 2005 (and after 2011), individuals with felony convictions were permanently disenfranchised unless they submitted an application to the governor. Executive Order 42 eliminated this requirement, instead re-enfranchising automatically upon being discharged from their sentence. Although all formerly disenfranchised individuals were re-enfranchised upon the signing of the Executive Order, not all re-enfranchised individuals were informed of their change in status. @Meredith2015 leverages differences in notification to test whether the notification increased turnout, finding a strong positive effect. @Meredith2013, however, examines states where so-called notification laws went into effect. Although rules about eligibility did not change in these states, new policies required Departments of Corrections to notify formerly disenfranchised individuals of their re-instated voting rights. @Meredith2013 finds no effect on turnout from notification in the absence of eligibility changes.

@Gerber2015 built on the quasi-experimental design of @Meredith2015 and conducted a field experiment in Connecticut in advance of the 2012 presidential election. In their field experiment, some formerly convicted (but eligible) residents were reminded of their eligibility; others were not. Like @Meredith2015, they find evidence that these reminders successfully increased turnout. "Whatever the participatory consequences of incarceration," they conclude, "they are not in large part impossible to overcome" (924). Even if incarceration does not decrease individuals' propensity to vote, there is reason to believe that reminding formerly incarcerated individuals of their rights increases participation.

@Meredith2015 and @Gerber2015 provide valuable insight into mechanisms that can increase the turnout of formerly disenfranchised individuals. Nevertheless, the research to date has examined the effect of mail notification on post-supervision turnout. In the case of New York State, formerly incarcerated individuals were being informed of their voting rights through the mail when they finished their sentence prior to the executive order. This study tests, therefore, not the efficacy of notification laws but rather whether in-person restoration prior to discharge increases turnout above-and-beyond any increase associated with mail notification.

There is reason to believe that some of the negative treatment effects of incarceration arise from the very nature of person-to-person relationships in the criminal justice system. @Lerman2014 argues that contact with the criminal justice system can uniquely structure democratic participation more than other types of government contact. "It may also be," they write, "that social benefits [arising from non-criminal justice contact with the state] are less visible to citizens because they often occur through the mail, rather than through personal contact with government agencies or officials, making it easier to disconnect social benefits from government and the political system" (93). Because Executive Order 181 operates *not* through the mail but rather through "personal contact with government... officials," its effects may differ from the mail notification programs. The social benefits (restoration of voting rights) might be communicated more effectively, thus increasing turnout further.

Of course, reliance on a human-mediated system to deliver social benefits poses potential practical and scientific complications. A notification sent in the mail is a binary treatment: the letter was either sent or not sent. Communicating the restoration of voting rights through human interaction means that the treatment might vary by individuals: if parolees have a strong relationship with their parole officer, they may be strongly encouraged to register and participate. On the other hand, if parole officers are biased against their parolees (due either to racial hostility, personal dislike, or expected partisan affiliation) they may not provide any effective treatment at all. Thus, while the communication of voting rights restoration in the context of a personal relationship has the potential to deliver much larger benefits, it also opens the door to potential discrepancies in treatment.

I expect that restoring voting rights while an individual is still on parole increases that individual's later propensity to vote substantially. This increase will operate through two mechanisms: as the previous literature has established, notification of voting rights can increase turnout, and being informed by a parole officer likely leaves a former parolee more sure of her voting rights than a letter in the mail. Secondly, the interpersonal nature of the parolee-parole officer relationship is expected to more effectively repair (some of) the damage done to the former parolees interpretation of the state. The individual, in-person invitation to rejoin the body politic was likely effective.

@@

Following Executive Order 181, the Department of Corrections and Community Supervision began indicating on their online Parolee Lookup Tool whether a parolee had her voting rights restored. By using the identification number provided from the parolee public records request and this website, I was able to identify individuals who had their voting rights restored.^[Not all parolees listed in the public records request data are included in the lookup tool. Roughly 1 percent of former parolees are not in the online lookup.] There were `r comma(readRDS("./temp/how_many_restored.rds"))` individuals who were discharged from parole before the registration deadline whose rights were restored while still under supervision.

@@

In this paper, I exploit randomness in discharge date from parole in New York State to determine the efficacy of the policy change. All individuals discharged from parole after the start of the executive order are assigned to the "treatment" group; individuals discharged earlier are assigned to the control group. When I limit the pool of formerly incarcerated individuals to those discharged from parole in 2017 and 2018, the control and treatment voters look virtually identical. The assignment to treatment, therefore, is as nearly random as can be hoped for in the context of a natural experiment.

The analysis is complicated, however, by the fact that not all individuals discharged from parole after the policy change --- that is to say, assigned to the treatment group --- were actually "treated" by having their voting rights restored while on parole in a meeting with their parole officer. In other words, our natural experiment has "imperfect compliance." Noncitizens, for instance, were not eligible for rights "restoration" while on parole because in New York State they were not allowed to vote even prior to their conviction. Because not all individuals assigned to the treatment group were in fact treated, simply comparing turnout between the groups discharged on either side of the cutpoint reveals not the treatment effect, but rather the intend-to-treat effect. Put differently, it allows us to estimate how much the executive order increased turnout among former parolees as a group --- including those ineligible to vote --- but not the individual-level effect of the policy change on propensity to vote.

Although we can identify the individuals in the assigned-to-treatment group who do not comply with their treatment status, we do not know who among the assigned-to-control group would have gone untreated if they had been assigned to the other group. We cannot identify, for instance, noncitizens who were discharged from parole before the policy change. In an ideal world, we would limit both the treatment *and* the control group to individuals who were eligible to have their rights restored while on parole; we do not live in such a world. Worse yet, there is obviously a relationship between compliance (rights restoration) and the dependent variable of interest: noncitizens are, of course, both ineligible to have voting rights bestowed and cannot vote.

The standard way of dealing with imperfect compliance in an experimental setting is to leverage the (random) assignment to treatment as an instrument for the (nonrandom) actual compliance [@Angrist1996]. Such an approach has been widely used in the context of voter turnout [e.g. @Ansolabehere1999; @Gerber2000; @Milligan2004; @Lassen2004; @Sondheimer2010]. @Gerber2000 provides a helpful example of why this two-step process is necessary. The authors test the efficacy of phone calls on increasing turnout. One group is assigned to be the control; the authors do not attempt to call them. Another set of voters receive a phone call, but only some of them are actually contacted, or "treated" with the call. The authors could simply test to see whether those contacted by phone turned out at higher rates --- but what if high propensity voters are more likely to answer the phone? Then "treatment" would simply be identifying a group of individuals with higher vote propensity than the control group without actually speaking to the efficacy of the treatment.

The same process is likely at play here. The control group includes individuals that *would have* had their rights restored at an in-person meeting if they had been discharged later; it also includes individuals who would not have received the treatment thanks to their citizenship status. By comparing turnout among the actually treated to the control group, we may simply be comparing the turnout of a group of *citizens* to the turnout of a group that includes both citizens *and* noncitizens. Although a positive "treatment effect" is likely in this context, it would be unable to speak to causality.

@@

Using the natural experiment with imperfect compliance setup, I begin by estimating the intend-to-treat effect of the policy change. This approach simply asks whether, as a whole, turnout was higher among individuals discharged after the policy change than those discharged earlier. I use turnout in the 2018 election as the dependent variable in these logistic regressions.

After estimating the intend-to-treat effect of the policy change, I implement the instrumental variables approach to estimate the individual-level effect of in-person voting rights restoration on subsequent turnout. As discussed above, the human nature of the treatment allows for discrepancies in how the treatment is administered. To explore this potential, I also explore whether the treatment had different effects for different subgroups of the formerly incarcerated population. I finally explore whether the treatment effects were larger for individuals who spent more time on parole after having their rights restored.

@@

```
\input{"../../temp/table3.tex"}
\end{singlespace}
The inclusion of time controls in Model 2 in Table \ref{tab:to-18-logit-short} leaves the AIC unchanged. A Chi-squared test confirms that the model is not improved when controls for time are included. When we look only at individuals recently discharged from parole, the length of time an individual has been off parole is not associated with his propensity to vote.^[The Supplemental Information provides further corroboration that being discharged from parole in the months before an election is uncorrelated with propensity to vote by exploring turnout rates in the 2016 presidential election.] This indicates that using individuals released in 2017, and 2018 prior to the executive order, as a control group is feasible.

Although it perhaps seems counterintuitive to throw out many of our observations, ignoring all individuals last discharged from parole prior to 2017 give us an important asset: it allows us to conceptualize Executive Order 181 as a natural experiment. Whether an individual was discharged from parole before or after May 18^th^, 2018, is akin to a randomly assigned treatment. Individuals who were discharged from parole prior to the implementation of the executive order are part of the "control group", while individuals discharged after are assigned "treatment" by the policy change. Because Table \ref{tab:to-18-logit-short} shows that discharge date from parole is uncorrelated in the short term with an individual's propensity to vote, any observed difference in turnout between individuals discharged before and after late May can therefore be attributed to the executive order.

>>>>

---
title: "Welcome Home --- Now Vote!"
subtitle: "Voting Rights Restoration and Post-Supervision Participation"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
  \linebreak
  \linebreak
  \linebreak
  \linebreak
  Word Count: 9,553
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "../welcome_home_vote.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
    - \usepackage{setspace}
abstract: |
  This paper presents causal estimates of the effect of voting rights restoration prior to discharge from parole on post-supervision participation. In 2018, New York State began restoring voting rights to parolees, after previously restoring voting rights only at the completion of parole. By leveraging randomness in parole discharge date, I interrogate whether restoring voting rights to parolees increases their post-supervision propensity to cast a ballot. I demonstrate that in-person rights restoration prior to parole discharge significantly increased turnout. This group-level effect, however, masks race-specific effects. Although rights restoration prior to parole discharge effectively doubled turnout among white former parolees, it had no measurable effect on the turnout of non-white former parolees. This raises serious questions about how rights restoration programs are implemented, and how incarceration might differently structure Black Americans’ view of the democratic process.
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\pagebreak
\doublespacing
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "../..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(data.table)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

In all but two states (Maine and Vermont), felony disenfranchisement laws mean that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. Iowa stands alone at the other end of the spectrum, where felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor [@bcj_laws]. This variation in laws arises from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights "for participation in rebellion, or other crime." The definition of "other crime," left so vague in the Constitution, is now generally used by states to disenfranchise citizens for any felony offense. The Supreme Court, in cases such as *Richardson v. Ramirez* (1974), has upheld this practice. Collectively, these laws disenfranchise as many as 4.7 million American citizens. Of these, the majority are no longer incarcerated [@sentencing_2016].^[The figures reported in @sentencing_2016 have been adjusted to reflect the impact of Amendment 4 in Florida.]

There is some evidence that incarceration continues to structure political participation even after an individual is no longer legally disenfranchised. As previous literature has established, interactions with the criminal justice system leaves residents less likely to vote in the future [@White2019; but see @Gerber2017]. As @Burch2011 and others have shown, moreover, turnout rates among the formerly incarcerated are extremely low. Formal disenfranchisement policy, the literature has made clear, is just one piece of an interlocking system that serves to disenfranchise minority and marginalized voters. The incarcerated population is drawn from a pool of individuals unlikely to vote even prior to their incarceration. To address only the formal laws contributing to disenfranchisement without also interrogating efforts to boost post-supervision participation risks leaving much of the system of effective disenfranchisement undisturbed. New York State offers us the opportunity to test how the timing and method of the re-instation of voting rights structures post-supervision participation.

@@

In the spring of 2018,^[Although the executive order was signed on April 18^th^, it did not go into effect until May 18^th^.] Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. Such a move was of course good for individuals who were still on parole on election day --- they would have been disenfranchised otherwise. The change in policy is also beneficial for felony probationers. Despite the fact that probationers do not formally lose their voting rights, there is evidence that confusion around the law contributes to *de facto* disenfranchisement among probationers [@Drucker2005]. The executive order is a promising step: by changing the policy to allow all New York citizens living in their communities to cast a ballot, it has the potential to both re-enfranchise the nearly 30,000 New Yorkers on parole living in the community and to clarify the rules about who is eligible to vote.

The executive order may also have increased the political participation of *formerly* disenfranchised individuals. Prior to the policy change, formerly incarcerated individuals had their voting rights restored automatically upon the completion of their parole term. New York's correction code provides no more guidance other than that "upon a person's discharge from community supervision, the department shall notify such person of his or her right to vote and provide such person with a form of application for voter registration" (Section 75 of the New York State Correction Law). Shortly after the implementation of the executive order, Acting Deputy Commissioner for Community Supervision Ana Enright sent a memorandum to New York State parole officers detailing the Department of Corrections' new approach.^[Link: https://nyassembly.gov/member_files/139/webdocs/82103.pdf] The memorandum directs all parole officers to present parolees with voter registration forms and to explain their purpose. Parole officers are instructed to offer any assistance needed, including help filling out the registration form. In addition to these directives, the memorandum communicates that voter registration is to receive "high priority attention," and it separately calls the program "a **priority** initiative" [emphasis in the original]. Thus, the executive order demands not only that re-enfranchised individuals receive in-person notification of their voting rights, but also that parole officers prioritize their registration.

In the analysis below, I examine the effect of Executive Order 181 on individuals who finished parole before October 10^th^, 2018 (the registration deadline for 2018). I exclude individuals who were still on parole and could only vote because of the policy change in order to unpack how the mechanisms of rights restoration can shape turnout.

By examining the effect of rights restoration in the context of a personal relationship with a parole officer against a status quo in which formerly incarcerated individuals were informed of their eligibility through the mail, Executive Order 181 offers us the opportunity to build on existing research. Do these human interactions repair damage done to the formerly incarcerated individual's relationship to the state? Or do parole officers exhibit biases against their stewards, making them ineffective or uneven conduits for restoration?

@@

Much of this literature establishes that formerly incarcerated individuals rarely vote, even when they are no longer formally barred from doing so. The causal effect of incarceration on participation is the subject of some debate within the field. Individuals who go to prison share many characteristics with lower propensity voters generally. Less educated citizens, for instance, turnout at low rates whether they have been to prison or not. In an attempt to disentangle sociodemographic characteristics from the experience of imprisonment, @Gerber2017 uses administrative data from Pennsylvania to estimate turnout rates prior to and after incarceration. They argue that the majority of the low turnout observed among formerly incarcerated individuals can be explained by observable characteristics, concluding that "it appears that spending time in prison does not have large negative effects on subsequent participation" (1144).

@@

Regardless of the precise mechanism, the low turnout among formerly incarcerated individuals is cause for concern, particularly given the racialized aspects of the criminal justice system. A criminal justice system that inflicts serious consequences on a population that has relatively little political voice is problematic. Whether or not incarceration *causes* low turnout, the state has a unique opportunity to craft policies that will impact individuals under formal supervision. Even if incarceration does not lead to lower turnout, policies targeting individuals caught up in the criminal justice system might still be effective at increasing their turnout.

Formerly convicted individuals are very often confused about their eligibility to vote [@Drucker2005; @locked_out]. Some research indicates that dispelling misinformation boosts turnout among the formerly disenfranchised. @Meredith2015 examines the impact of ending permanent disenfranchisement in Iowa. Prior to 2005 (and after 2011), individuals with felony convictions were permanently disenfranchised unless they submitted an application to the governor. Executive Order 42 eliminated this requirement, instead re-enfranchising individuals automatically upon sentence completion. Although all formerly disenfranchised individuals were re-enfranchised upon the signing of the Executive Order, not all re-enfranchised individuals were informed of their change in status. @Meredith2015 leverages differences in notification to test whether the notification increased turnout, finding a strong positive effect. @Meredith2013, on the other hand, examines states where so-called notification laws went into effect. Although rules about eligibility did not change in these states, new policies required Departments of Corrections to notify formerly disenfranchised individuals of their re-instated voting rights. @Meredith2013 finds no effect on turnout from notification in the absence of eligibility changes.

@Gerber2015 built on the quasi-experimental design of @Meredith2015, conducting a field experiment in Connecticut in advance of the 2012 presidential election. In their field experiment, some formerly convicted (but eligible) residents were reminded of their eligibility; others were not. Like @Meredith2015, they find evidence that these reminders successfully increased turnout. "Whatever the participatory consequences of incarceration," they conclude, "they are not in large part impossible to overcome" (924). Even if incarceration does not decrease individuals' propensity to vote, there is reason to believe that reminding formerly incarcerated individuals of their rights increases participation.

@Meredith2015 and @Gerber2015 provide valuable insight into mechanisms that can increase the turnout of formerly disenfranchised individuals. Nevertheless, the research to date has examined the effect of mail notification on post-supervision turnout. In the case of New York State, formerly incarcerated individuals were already informed of their voting rights through the mail when they finished their sentence prior to the executive order. In other words, the treatment identified in past research is here incorporated into the status quo, or control, condition. This study therefore tests not the efficacy of notification laws but rather whether in-person restoration prior to discharge increases turnout above-and-beyond any increase associated with mail notification.

There is reason to believe that some of the negative treatment effects of incarceration arise from the very nature of person-to-person relationships in the criminal justice system. @Lerman2014 argues that contact with the criminal justice system can uniquely structure democratic participation more than other types of government contact. "It may also be," they write, "that social benefits [arising from non-criminal justice contact with the state] are less visible to citizens because they often occur through the mail, rather than through personal contact with government agencies or officials, making it easier to disconnect social benefits from government and the political system" (93). Because Executive Order 181 operates *not* through the mail but rather through "personal contact with government... officials," its effects may differ from the mail notification programs. The social benefits (restoration of voting rights) might be communicated more effectively and in ways that improve the formerly incarcerated individuals' opinions of their democratic state, thus increasing turnout further.

Of course, reliance on a human-mediated system to deliver social benefits poses potential practical and scientific complications. A notification sent in the mail is a binary treatment: the letter was either sent or not sent. Communicating the restoration of voting rights through human interaction means that the treatment might vary by individuals: if parolees have a warm relationship with their parole officer, they may be strongly encouraged to register and participate. On the other hand, if parole officers are biased against their parolees (due either to racial hostility, personal dislike, or expected partisan affiliation) they may not provide any effective treatment at all. Thus, while the communication of voting rights restoration in the context of a personal relationship has the potential to deliver much larger benefits, it also opens the door to potential discrepancies in treatment.

I expect that restoring voting rights while an individual is still on parole increases that individual's later propensity to vote substantially. This increase will operate through two mechanisms: as the previous literature has established, notification of voting rights can increase turnout, and being informed by a parole officer likely leaves a former parolee more sure of her voting rights than a letter in the mail. Secondly, the interpersonal nature of the parolee-parole officer relationship is expected to more effectively repair (some of) the damage done to the former parolees interpretation of the state. Individual, in-person invitations to rejoin the body politic are likely effective.

@@

Following Executive Order 181, the Department of Corrections and Community Supervision began indicating on their online Parolee Lookup Tool whether a parolee had her voting rights restored. By using the identification number provided from the parolee public records request and this website, I was able to identify individuals who had their voting rights restored.^[Not all parolees listed in the public records request data are included in the lookup tool. Roughly 1 percent of former parolees are not in the online lookup.] There were `r comma(readRDS("./temp/how_many_restored.rds"))` individuals who were discharged from parole before the registration deadline whose rights were restored while still under supervision. Roughly 1,200 individuals who were discharged between the effective date of the executive order and the registration deadline did not have voting rights restored, due largely to their status as noncitizens.

@@

In this paper, I exploit randomness in discharge date from parole in New York State to determine the efficacy of the policy change. All individuals discharged from parole after the effective date of the executive order are assigned to the treatment group; individuals discharged earlier are assigned to the control group. When I limit the pool of formerly incarcerated individuals to those discharged from parole in 2017 and 2018, the control and treatment voters look virtually identical. The assignment to treatment, therefore, is as nearly random as can be hoped for in the context of a natural experiment.

The analysis is complicated, however, by the fact that not all individuals discharged from parole after the policy change --- that is to say, assigned to the treatment group --- were actually treated by having their voting rights restored while on parole in a meeting with their parole officer. In other words, our natural experiment has "imperfect compliance" [see, for instance, @Angrist1996]. Noncitizens, for instance, were not eligible for rights "restoration" while on parole because in New York State they were not allowed to vote even prior to their conviction. Because not all individuals assigned to the treatment group were in fact treated, simply comparing turnout between the groups discharged on either side of the cutpoint reveals not the treatment effect, but rather the intent-to-treat effect. Put differently, it allows us to estimate how much the executive order increased turnout among former parolees as a group --- including those ineligible to vote --- but not the individual-level effect of the policy change on propensity to vote.

Although we can identify the individuals in the assigned-to-treatment group who do not comply with their treatment status, we do not know who among the assigned-to-control group would have gone untreated if they had been assigned to the other group. We cannot identify, for instance, noncitizens who were discharged from parole before the policy change. In an ideal world, we would limit both the assigned-to-treatment *and* the assigned-to-control groups to voting-eligible individuals; we do not live in such a world. Worse yet, there is an obvious relationship between compliance (rights restoration) and the dependent variable of interest: noncitizens are both ineligible to have voting rights bestowed and cannot vote.

The standard way of dealing with imperfect compliance in an experimental setting is to leverage the (random) assignment to treatment as an instrument for the (nonrandom) actual compliance [@Angrist1996]. Such an approach has been widely used in the context of voter turnout [e.g. @Ansolabehere1999; @Gerber2000; @Milligan2004; @Lassen2004; @Sondheimer2010]. @Gerber2000 provides a helpful example of why this two-step process is necessary. The authors test the efficacy of phone calls on increasing turnout. One group is assigned to be the control; the authors do not attempt to call them. Another set of voters receive a phone call, but only some of them are actually contacted, or "treated" with the call. The authors could simply test to see whether those contacted by phone turned out at higher rates --- but what if high propensity voters are more likely to answer the phone? Then "treatment" would simply identify a group of individuals with higher vote propensity than the control group without actually speaking to the efficacy of the treatment.

The same process is likely at play here. The control group includes individuals that *would have* had their rights restored at an in-person meeting if they had been discharged later; it also includes individuals who would not have received the treatment thanks to their citizenship status. By comparing turnout among the actually treated to the control group, we may simply be comparing the turnout of a group of *citizens* to the turnout of a group that includes both citizens *and* noncitizens. Although a positive "treatment effect" is likely in this context, it would be unable to speak to causality. The instrumental variables approach allows us to mitigate this problem.

@@

Using the natural experiment with imperfect compliance setup, I begin by estimating the intent-to-treat effect of the policy change. This approach simply asks whether, as a whole, turnout was higher among individuals discharged after the policy change than those discharged earlier. I use turnout in the 2018 election as the dependent variable in these logistic regressions.

After estimating the intent-to-treat effect of the policy change, I implement the instrumental variables approach to estimate the individual-level effect of in-person voting rights restoration on subsequent turnout. As discussed above, the human nature of the treatment allows for discrepancies in how the treatment is administered. To explore this potential, I also test whether the treatment had different effects for different subgroups of the formerly incarcerated population. I finally ask whether the treatment effects were larger for individuals who spent more time on parole after having their rights restored.

This paper is primarily interested in the effect of in-person rights restoration prior to parole discharge on ultimate political representation. To that end I use turnout --— not registration —-- as the dependent variable of interest throughout. The Supplemental Information demonstrates that all treatment effects discussed here (including for subgroups within the population) also hold when the dependent variable is registration, not turnout.

@@

```
\input{"../../temp/table3.tex"}
\end{singlespace}
The inclusion of time controls in Model 2 in Table \ref{tab:to-18-logit-short} leaves the AIC unchanged. A Chi-squared test confirms that the model is not improved when controls for time are included. When we look only at individuals recently discharged from parole, the length of time an individual has been off parole is not associated with his propensity to vote.^[The Supplemental Information provides further corroboration that being discharged from parole in the months before an election is uncorrelated with propensity to vote by exploring turnout rates in the 2016 presidential election.] This indicates that using individuals released in 2017, and 2018 prior to the executive order, as a control group is feasible. Because discharge date from parole is uncorrelated in the short term with an individual's propensity to vote in the short-run, any observed difference in turnout between individuals discharged before and after late May can therefore be attributed to the executive order.

To further demonstrate the usefulness of discharge date as an intent-to-treat indicator, Table \ref{tab:demo-rd} shows the demographic characteristics of individuals in the control group (those discharged between January 1st, 2017, and May 17th, 2018) and the intent-to-treat group (those discharged between May 18th and October 12th, 2018). With the exception of age, the control and intent-to-treat groups are statistically indistinguishable from one another, further demonstrating the validity of the natural experiment conceptualization. The control group is, on average, slightly older, which Table \ref{tab:to-18-logit-short} indicates makes them more likely to vote. To the extent our control group perhaps has a marginally higher propensity to vote than our intent-to-treat group our setup is (slightly) biased against finding a significant increase due to the executive order.
\begin{singlespace}
```{r rd-demo-chunk, echo = F, warning = F}
t <- readRDS("./temp/demos_rd_time.rds")
j <- knitr::kable(t, booktabs = T, caption = "(\\#tab:demo-rd) Demographics of \"Control\" and \"intent-to-Treat\" Groups", linesep = "",
                  align = c("l", "c", "c"), "latex") %>% 
  kable_styling(latex_options = c("HOLD_position"),
                full_width = F)
